\Title

### 2013-02-13 synmut.tex
types and edits in figure captions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aftsyn}, this is a simple consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular locus and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course the frequencies of all synonymous and
nonsynonymous mutations observed \env, \shankaregion, in patient
p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (patients whose viral population was structured
were excluded from the analysis; see methods and \figurename~S\PCApat).  The
former data set from is restricted to the \shankaregion{} region of \env, while
the data from \citet{bunnik_autologous_2008} covers the majority of \env.
Considering all mutations in a frequency interval around $\nu_0$ at some time
$t$, we calculate the fraction that is found at frequency 1, at frequency 0, or
at intermediate frequency at later times $t+\Delta t$. Plotting these fixed,
lost, and polymorphic fraction against the time interval $\Delta t$, we see that
most synonymous mutations segregate for roughly one year and are lost much more
frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
fixation versus extinction of synonymous mutations is shown as a function of the
initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). Restricted to the
region \shankaregion, we find that $\pfix$ of synonymous mutations is far below
the neutral expectation.  Outside of \shankaregion, using data from
\citet{bunnik_autologous_2008} only, no such reduction in $\pfix$ is found.
Restricted to the \shankaregion area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will
come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, \shankaregion, from patient
1~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms observed at a certain
frequency since the latter have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
reached a high frequency by chance (genetic drift or hitchhiking), and
we expect that selection drives it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env{} that includes the hypervariable regions are
deleterious, while outside this regions polymorphisms are mostly roughly
neutral. Note that this does not imply that all synonymous mutations in this
region are neutral -- only those mutations observed at high frequencies, which
have been experiencing selection for some time already, tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows the time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in \shankaregion is reduced relative to the neutral
expectation indicated by the diagonal line. This suppression is not
observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

The propensity of nucleotides in the HIV genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems. We partition all synonymous alleles observed at
intermediate frequencies above 15\% depending on their final destiny (fixation
or extinction). Subsequently, we align our sequences to the reference NL4-3
strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
reactivities. As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $P\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between the those that fix and
those that get lost. Note that this analysis will be sensitive only at position
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

To test the hypothesis that mutations in \shankaregion{} are lost because they
break stems in the conserved stretches between the variable loops, we consider
mutations in variable loops and conserved parts separately. The biggest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is consistent with important stem structures in conservered
regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among synonymous mutations
that fix, mutations that reach frequencies above 15\%, but
are susequently lost, and random mutations (all categories are restricted to the
regions V1-V5$\pm 100$bp).
Fixed mutations tend to have higher higher SHAPE reactivities, corresponding to
less base pairing.
Mutations that are never observered show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous mutations in
\shankaregion{} separately for variable loops and the connecting conserved regions that
harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous the
fixation probability and diversity observed in patients. Parameters are chosen
from uniform prior distributions as indicated by the red box (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

@@

The above simulation show that hitchhiking can explain the obervation of
deleterious mutations that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV
infection, nonsynonymous mutations at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
test two possible mechanisms that are biologically sound and could explain
the transient rise of nonsynonymous mutations: time-dependent selection and
within-epitope competition.

@@

While classical population genetics assumes that the dominant stochastic force
is genetic drift, i.e. nonheritable fluctuations in offspring number, our
results show that stochasticity due to linked selection is much more important.
Such fluctuations have been termed \emph{genetic draft} by
\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
such as HIV has been characterized in \citep{neher_genetic_2011}. Importantly,
large population sizes are compatible with low diversity and fast coalescence
when draft dominates over drift.

Contrary to na\"ive expectations, the adaptive escape mutations do not seem to
be unconditionally beneficial. Otherwise we would observe almost sure fixation
of nonsynonymous mutations once they reach intermediate frequencies. Instead, we
find that the fixation probability of nonsynonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation.
Similar to synonymous mutations, the majority of nonsynonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield a
more neutral-like averaged fixation probability. We consider this possibility
inplausible since the amino acid sequence outside the variable loops is much
more conserved that the synonymous positions, suggesting that the majority of
the nonsynonymous mutations is much more deleterious.

Alternatively, the lack of fixation could be due to time dependent environment
through an immune system that is catching up, or competition between mutations
that mediate escape within the same epitope. We explore both of these
possibilities and find that both produce the desired effect. Furthermore, there
is experimental evidence in support of both of these hypotheses. Serum from HIV
infected individuals tyically neutralizes the virus that dominated the
population a few (3-6) month ago \citep{richman_rapid_2003}. This suggests that
escape mutations cease to be beneficial after a few months and might revert if
they come with a fitness cost. Deep sequencing of regions of \env{} after
antibody escape have revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them do not provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large population size that explores all necessary point
mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
in the context of preexisting drug resistance mutations
\citep{boltz_ultrasensitive_2012}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The sequences were translated and the resulting amino acid sequences aligned
using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
sequences separately for each patient. Within each patient, the consensus
nucleotide sequence at the first time point was used to classify alleles as
``ancestral'' or ``derived'' at all sites. Sites that include large
frequencies of gaps were excluded from the analysis to avoid artefactual
substitutions due to alignment errors. Allele frequencies at different time
points were extracted from the multiple sequence alignment.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aftsyn}, this is a simple consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular locus and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course the frequencies of all synonymous and
nonsynonymous mutations observed \env, \shankaregion, in patient
p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (patients whose viral population was structured
were excluded from the analysis; see methods and \figurename~S\PCApat).  The
former data set from is restricted to the \shankaregion{} region of \env, while
the data from \citet{bunnik_autologous_2008} covers the majority of \env.
Considering all mutations in a frequency interval around $\nu_0$ at some time
$t$, we calculate the fraction that is found at frequency 1, at frequency 0, or
at intermediate frequency at later times $t+\Delta t$. Plotting these fixed,
lost, and polymorphic fraction against the time interval $\Delta t$, we see that
most synonymous mutations segregate for roughly one year and are lost much more
frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
fixation versus extinction of synonymous mutations is shown as a function of the
initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). Restricted to the
region \shankaregion, we find that $\pfix$ of synonymous mutations is far below
the neutral expectation.  Outside of \shankaregion, using data from
\citet{bunnik_autologous_2008} only, no such reduction in $\pfix$ is found.
Restricted to the \shankaregion{} region, the sequence samples from
\citet{bunnik_autologous_2008} are fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will
come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, \shankaregion, from patient
10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
mutation is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms observed at a certain
frequency since the latter have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
reached a high frequency by chance (genetic drift or hitchhiking), and
we expect that selection drives it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env{} that includes the hypervariable regions are
deleterious, while outside this regions most polymorphisms are roughly
neutral. Note that this does not imply that all synonymous mutations in this
region are neutral -- only those mutations observed at high frequencies, which
have been experiencing selection for some time already, tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows the fraction of polymorphisms that have fixed (lower curves),
are lost (distances between upper curves and 1), or remain polymorphic a time
$\Delta t$ later for different initial allele frequencies (colors, indicated as
shaded area).
In each frequency interval, the fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) shows the fixation probability of derived synonymous
alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
often than expected for neutral mutations indicated by the diagonal line.
This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

The propensity of nucleotides in the HIV genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems. We partition all synonymous alleles observed at
intermediate frequencies above 15\% depending on their final destiny (fixation
or extinction). Subsequently, we align our sequences to the reference NL4-3
strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
reactivities. As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between the those that fix and
those that get lost. Note that this analysis will be sensitive only at position
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

To test the hypothesis that mutations in \shankaregion{} are lost because they
break stems in the conserved stretches between the variable loops, we consider
mutations in variable loops and conserved parts separately. The biggest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is consistent with important stem structures in conserved
regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among synonymous mutations
that fix, mutations that reach frequencies above 15\%, but
are susequently lost, and random mutations (all categories are restricted to the
regions V1-V5$\pm 100$bp).
Fixed mutations tend to have higher higher SHAPE reactivities, corresponding to
less base pairing.
Mutations that are never observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous mutations in
\shankaregion{} separately for variable loops and the connecting conserved regions that
harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous the
fixation probability and diversity observed in patients. Parameters are chosen
from uniform prior distributions as indicated by the red box (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

@@

The above simulation show that hitchhiking can explain the observation of
deleterious mutations that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV
infection, nonsynonymous mutations at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
test two possible mechanisms that are biologically sound and could explain
the transient rise of nonsynonymous mutations: time-dependent selection and
within-epitope competition.

@@

While classical population genetics assumes that the dominant stochastic force
is genetic drift, i.e. non-heritable fluctuations in offspring number, our
results show that stochasticity due to linked selection is much more important.
Such fluctuations have been termed \emph{genetic draft} by
\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
such as HIV has been characterized in \citep{neher_genetic_2011}. Importantly,
large population sizes are compatible with low diversity and fast coalescence
when draft dominates over drift.

Contrary to na\"ive expectations, the adaptive escape mutations do not seem to
be unconditionally beneficial. Otherwise we would observe almost sure fixation
of nonsynonymous mutations once they reach intermediate frequencies. Instead, we
find that the fixation probability of nonsynonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation.
Similar to synonymous mutations, the majority of nonsynonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield a
more neutral-like averaged fixation probability. We consider this possibility
implausible since the amino acid sequence outside the variable loops is much
more conserved that the synonymous positions, suggesting that the majority of
the nonsynonymous mutations is much more deleterious.

Alternatively, the lack of fixation could be due to time dependent environment
through an immune system that is catching up, or competition between mutations
that mediate escape within the same epitope. We explore both of these
possibilities and find that both produce the desired effect. Furthermore, there
is experimental evidence in support of both of these hypotheses. Serum from HIV
infected individuals typically neutralizes the virus that dominated the
population a few (3-6) month ago \citep{richman_rapid_2003}. This suggests that
escape mutations cease to be beneficial after a few months and might revert if
they come with a fitness cost. Deep sequencing of regions of \env{} after
antibody escape have revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them do not provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large population size that explores all necessary point
mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
in the context of preexisting drug resistance mutations
\citep{boltz_ultrasensitive_2012}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The sequences were translated and the resulting amino acid sequences aligned
using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
sequences separately for each patient. Within each patient, the consensus
nucleotide sequence at the first time point was used to classify alleles as
``ancestral'' or ``derived'' at all sites. Sites that include large
frequencies of gaps were excluded from the analysis to avoid artifactual
substitutions due to alignment errors. Allele frequencies at different time
points were extracted from the multiple sequence alignment.
